Fail Faster

### 2025-03-22 paper/paper.tex
blah

In \camlname, our strategies include type-derived generators for BSTs and STLC terms, 
following the approach described in \secref{subsection:type-derived}; two custom 
BST generators—--one that builds a tree incrementally by repeatedly inserting 
values into an initially empty structure, and another that constructs the 
entire tree in a single pass by generating keys, values, and subtrees at each step; 
a boolean list generator that mirrors the single-pass BST strategy; and a 
generator for STLC terms that is correct by construction (i.e., it produces 
only well-typed terms).

@@

For each strategy, we compare three treatments: our \bq baseline; 
a \camlname version using \bq's randomness 
library, \texttt{splittable\_random}; and a
\camlname version using \csplitmix. For each treatment, we measure the
time to generate a value---i.e., a BST, a STLC term, or a boolean list---using a random seed, over a 
variety of different generation sizes (10, 100, 1,000, 10,000).\footnote{The way size is used in a generator is a domain-specific implementation detail---in a list generator, size might correspond to the desired length of the list, whereas in a tree generator it refer to number of nodes, number of leaves, depth of tree, etc. Regardless of how size is encoded, our goal is to show that performance trends scale.}
We run each treatment for 5 seconds---enough time for several thousand generations---and compute the average generation time of all values produced in that interval.

@@

The boolean list generator experiences $2.77-7.47\times$ speedups with
staging, and its performance changes very little when
combined with \csplitmix ($2.77-7.57\times$). This is likely because sampling
booleans is cheap enough that the overhead of crossing the FFI barrier is
comparable to that of generating values directly in OCaml.

To investigate the relationship between a generator's use of library 
abstractions and the performance benefits it gains from staging, 
we focused on the \texttt{bind} function---a core component of monadic 
generator libraries, frequently invoked both directly and indirectly 
through other combinators. We instrumented \texttt{bind} to count how many 
times it was called during a generator's execution. Our 
hypothesis was that generators which make heavier use of 
library abstractions like \texttt{bind} would benefit more from staging.

We evaluated this hypothesis using four benchmarks with widely varying 
speedups: insertion-based BSTs, single-pass BSTs, boolean lists, and 
well-typed STLC terms. For each generator, we ran 10,000 trials, 
using a fixed generation size of 100, and recorded the average number 
of \texttt{bind} calls. We then compared these usage statistics to the 
observed speedups.

As shown in \figref{fig:binds}, the results reveal a clear linear 
correlation between the number of \texttt{bind} calls and the speedup 
achieved through staging. This suggests that generators that 
rely more heavily on the library's abstractions tend to 
benefit more from staging optimizations. \tr{TALK ABOUT RANDOMNESS HERE.}

@@

>>>>

In \camlname, our strategies include type-derived generators for BSTs and STLC terms, 
following the approach described in \secref{subsection:type-derived}; two custom 
BST generators---one that builds a tree incrementally by repeatedly inserting 
values into an initially empty structure, and another that constructs the 
entire tree in a single pass by generating keys, values, and subtrees at each step; 
a boolean list generator that mirrors the single-pass BST strategy; and a 
generator for STLC terms that is correct by construction (i.e., it produces 
only well-typed terms).

@@

For each strategy, we compare three treatments: our \bq baseline; 
a \camlname version using \bq's randomness 
library, \texttt{splittable\_random}; and a
\camlname version using \csplitmix. For each treatment, we measure the
time to generate a value (i.e., a BST, a STLC term, or a boolean list), using a random seed, varying generation sizes (10, 100, 1,000, 10,000).\footnote{The way size is used in a generator is a domain-specific implementation detail---in a list generator, size might correspond to the desired length of the list, whereas in a tree generator it refer to number of nodes, number of leaves, depth of tree, etc. Regardless of how size is encoded, our goal is to show that performance trends scale.}
We run each treatment for 5 seconds---enough for several thousand generations---and compute the average generation time of a value produced in that interval.

@@

The boolean list generator experiences $2.77-7.47\times$ speedups with
staging, and its performance changes minimally when
combined with \csplitmix ($2.77-7.57\times$). This is likely because sampling
booleans is cheap enough that the overhead of crossing the FFI barrier is
comparable to that of generating values directly in OCaml.

The variation in speedups raises a natural question: what 
determines how much a generator benefits from staging? One 
likely factor is its degree of reliance on generator library functions.
This is hard to quantify directly, since not all functions 
contribute equally to overhead. As a proxy, we count how many 
times a generator calls \texttt{bind}---the central function in 
monadic generator libraries, extensively used both directly and within 
other combinators.

We select four benchmarks with disparate speedups: insertion-based BSTs, single-pass BSTs, 
boolean lists, and well-typed STLC terms. For each generator, 
we run 10,000 trials using a fixed generation size of 100, and 
record the average number of \texttt{bind} calls. Results in \figref{fig:binds} show a clear linear 
correlation between the number of \texttt{bind} calls and the speedup 
achieved by staging, supporting the idea that generators which make heavier use of 
generator library functions tend to benefit more.

@@

Similarly, we investigate the impact of \csplitmix by tracking the number of calls to \texttt{next\_int64} in \bq's \texttt{splittable\_random} 
library, which samples the next random integer.
This, too, is an imperfect proxy, as the SplitMix algorithm's runtime does plenty of stuff that isn't 
sampling which can dominate its runtime---which is all faster in \csplitmix. However, it too validates our intuition that sampling-heavy generators achieve
greater benefit from \csplitmix than they do staging:
